### 2016-07-04 case-studies.tex
various edits to regain space

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rr}
isolation of MySQL may lead to anomalous executions as confirmed by
our experiments; with 32 concurrent writers performing single
increments, and one reader issuing a constant stream of reads, we
observed a maximum of 12 violations of monotonicity in 10 rounds of
operation. We subsequently instantiated our reasoning framework for an
{\sc sc} store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
formally reason about scenarios described in this section can be found
in the supplementary.} In particular, we were able to formally prove
that to preserve the invariant, writers need to execute at
\iso{Serializable} isolation, while readers can execute at \iso{Read
Uncommitted}. Repeating the experiments with this configuration led to
no violations, as expected. While executing all transactions at {\sc
ser} also prevents violations, it led to an additional 113\% increase
in the latency of writes on an average (32 concurrent writers; 5
rounds). We repeated the experiments with the same configuration on
Postgres, and found no violations. However, Postgres's isolation
levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
ways. We therefore formalized Postgres's guarantees and repeated
relevant parts of the proof to discover that \iso{Repeatable Read} is
in fact sufficient for writers - an observation we confirmed through
experiments. As usual, using {\sc ser} for all transactions also
suffices, but on Postgres it led to an additional 47\% increase in
write latency.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
re-implemented the counter as a op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident as we
considered an instantiation our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
In other words, if the reads chose the largest valued write effect,
instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}.
We implemented this version of the counter as an op-based RDT, which
uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
latency compared to the version that uses serializable increments.

\paragraph{Microblog}

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behavior is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, chosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. To determine appropriate isolation levels,
we then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expectected), while yeilding a 38\% better overall latency than {\sc
ser} configuration (96 concurrent clients; 5 rounds).  Since
Postgres's isolation levels are stronger than the corresponding ANSI
levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
deduce that the same choice of isolation levels also work on Postgres.
Indeed, experiments confirm this judgment.

As evident from our case studies, the there already exists a
significant diversity among the weak consistency and isolation
variants implemented on various real-world data stores. More such
variants have been developed in the recent years to meet the need for
high availability on geo-replicated stores. For example,
\emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
targeted for replicated stores. As more commercial weakly consistent
stores adopt various kinds of transactional semantics, like the
recently introduced \emph{lightweight transactions} of Cassandra, it
is reasonable to expect new isolation definitions to continue to be
proposed. It may be possible to carefully engineer a reasoning
framework for each combination of consistency and isolation, but such
a strategy would lead to multiple semantic definitions and proof
systems with no obvious way to compare and relate them. Having a
parameterized semantics and a proof framework built on the such
semantics allows us to support these variants as distinct, yet
comparable instantiations, as we demonstrate in our case studies.
%  GK: what does the following sentence mean?
%  As a consequence, we avoid the need for semantics re-engineering
%  every time an application is retargeted to a new platform.

>>>>

\paragraph{A video view counter.} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
  tools we developed to analyze Ruby can be found at the following
  anonymized url: \url{...}}. The application stores the video count
in a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default
\iso{Repeatable Read} isolation level of MySQL leads to anomalous
executions - with 32 concurrent writers performing single increments,
and one reader issuing a constant stream of reads, we observed 12
violations of monotonicity in 10 rounds. We subsequently instantiated
our reasoning framework for an {\sc sc} store with ANSI SQL isolation
levels, and spent 5 man hours determining appropriate isolation levels
needed to enforce the invariant\footnote{The representative \txnimp
  programs we used to formally reason about scenarios described in
  this section can be found in the supplementary.} In particular, we
were able to formally prove that to preserve the invariant, writers
need to execute at \iso{Serializable} isolation, while readers can
execute at \iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes on an average (32
concurrent writers; 5 rounds). We repeated the experiments with the
same configuration on Postgres, and found no violations. However,
Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. As usual, using {\sc
  ser} for all transactions also suffices, but on Postgres results in
an additional 47\% increase in write latency.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency levels, we
re-implemented the counter as an op-based \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
isolation levels. Keeping the consistency level at {\sc ec}, we then
repeated our experiments with committed reads and serializable
increments - a combination which was proved to preserve monotonicity
on an {\sc sc} store, but nonetheless observed monotonicity
violations. The reason for such violations became evident when
we instantiated our proof framework for an {\sc ec} store.
Under this instantiation, it is impossible to derive the
\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
that successive reads in a session witness monotonically increasing
state. However, assuming {\sc cc} store consistency for reads allowed
us to complete the proof, and recover the monotonicity invariant. In
fact, we were able to complete the proof even after relaxing the
serializability requirement for increments, just by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
In other words, if the reads chose the largest valued write effect,
instead of the latest write effect they witness, then counter
increments need not be serializable for reads to witness a
monotonically increasing count\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}.
We implemented this version of the counter as an op-based RDT, which
uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
latency compared to the version that uses serializable increments.
\vspace*{-3pt}
\paragraph{Microblog.} Our second set of experiments focused on a Twitter-like microblogging
site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
invariants that the application chooses to enforce is referential
integrity between microposts and authors. The application enforces
this invariant entirely in user space by relying on database-backed
transactions (this behavior is typical of Rails
applications~\cite{bailisferal}).  For instance, deletion of a user is
carried out in a transaction that also includes the deletion of
microposts authored by the user.  However, concurrent observers can
nonetheless witness violations of referential integrity. To measure
the likelihood of such anomalies, we first populated a Postgres
database with 1000 user accounts, each user with 50 microposts and 20
followers, chosen uniformly at random. We then constructed experiments
with 64 concurrent clients performing deletions and timeline reads of
randomly chosen users in a 1:7 ratio, and witnessed 250 violations of
referential integrity. To determine appropriate isolation levels, we
then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expectected), while yielding a 38\% reduction in latency than an {\sc
  ser} configuration (96 concurrent clients; 5 rounds).  Since
Postgres's isolation levels are stronger than the corresponding ANSI
levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
immediately deduce that the same choice of isolation levels also work
on Postgres.

Clearly, there exists a significant diversity among the weak
consistency and isolation variants implemented on various real-world
data stores. More such variants have been developed in recent years to
meet the need for high availability (\emph{e.g.}, \emph{Parallel
  Snapshot Isolation}~\cite{psi} (2011) or \emph{Non-Monotonic
  Snapshot Isolation}~\cite{nmsi} (2013)).  As more commercial
weakly-consistent stores adopt various kinds of transactional
semantics, like Cassandra's recently introduced \emph{lightweight
  transactions}, it is reasonable to expect new isolation
definitions to continue to be proposed.  While it may be possible to
carefully engineer a reasoning framework for each combination of
consistency and isolation, such a strategy would lead to multiple
semantic definitions and proof systems with no obvious way to compare
and relate them. As illustrated here, having a parameterized semantics
and a proof framework built on top of it allows us to support these
variants as distinct, yet comparable, instantiations.

%  GK: what does the following sentence mean?
%  As a consequence, we avoid the need for semantics re-engineering
%  every time an application is retargeted to a new platform.

### 2016-07-04 introduction.tex
various edits to regain space

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some of well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments.  Consequently, reasoning about weak isolation remains an
error prone endeavor, with major database vendors~\cite{postgresiso,
mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden squarely on the programmer to
determine application correctness.

@@

\noindent Our results thus provide the first (to the best of our
knowledge) mechanism that precisely relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

The remainder of the paper is organized as follows.  The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly consistent data stores.  We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:casestudies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

>>>>

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some of well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments.  Consequently, reasoning about weak isolation remains an
error prone endeavor, with major database vendors~\cite{postgresiso,
  mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden on the programmer to determine
application correctness.

@@

\noindent Our results provide the first (to the best of our
knowledge) mechanism that precisely relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

The remainder of the paper is organized as follows.  The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly consistent data stores. We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:case-studies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

### 2016-07-04 motivation.tex
various edits to regain space

\begin{figure}
\centering
$\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Wd1'$\rangle${
    if (C $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Wd2'$\rangle${
    if (C $\ge$ a2) {
      C := C - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\texttt{C}=\texttt{k-a1-a2}\}$

\caption{Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\end{figure}

@@

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-ex}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{Visibility and interference properties for different weak isolation levels can be
  captured by different instantiations of their happens-before relation.}
\label{fig:interference-ex}
\end{figure}

@@

\caption{$\visZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
  and dashed) isolation levels for the example in
  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\visZ$ relationship
  among all operations of `Wd2' and operations in `Wd1' that follow
  the arrow head. Letters R, W and C stand for read, write and commit,
  respectively.}
\label{fig:motiv-eg-1-hb}
\end{figure}

>>>>

\begin{figure}
\centering
$\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Wd1'$\rangle${
    if (C $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Wd2'$\rangle${
    if (C $\ge$ a2) {
      C := C - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\texttt{C}=\texttt{k-a1-a2}\}$
\caption{\small Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\vspace*{-10pt}
\end{figure}

@@

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-ex}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Visibility and interference properties for different weak isolation levels can be
  captured by different instantiations of their happens-before relation.}
\label{fig:interference-ex}
\end{figure}

@@

\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
  and dashed) isolation levels for the example in
  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
  among all operations of `Wd2' and operations in `Wd1' that follow
  the arrow head. Letters R, W and C stand for read, write and commit,
  respectively.}
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

### 2016-07-04 reasoning.tex
various edits to regain space

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
these definitions also require a \emph{stability} condition, which
requires pre- and post- conditions to hold despite any interference
from concurrent threads (captured by $R$). Stability can be predicated
on the assumption that interference preserves the trace invariant
$\I$. Formally:

@@

\noindent However, the assumption that interference preserves $\I$ (or,
dually, $\I$ withstands interference) needs to be justified
separately. We call this the stability requirement on $\I$:

@@

\begin{definition}[\bfseries Interleaved step and multi-step relations]
An interleaved step relation interleaves thread-local reductions with
interference from concurrent threads captured as the rely relation
($R$).  It is defined thus:

@@

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation thus:

@@

\noindent The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to an RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:

>>>>

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
these definitions also require a \emph{stability} condition, which
requires pre- and post- conditions to hold despite any interference
from concurrent threads (captured by $R$). Stability can be predicated
on the assumption that interference preserves the trace invariant
$\I$. Formally:\vspace*{-10pt}

@@

\noindent However, the assumption that interference preserves $\I$ (or,
dually, $\I$ withstands interference) needs to be justified
separately. We call this the stability requirement on $\I$:\vspace*{-10pt}

@@

\begin{definition}[\bfseries Interleaved step and multi-step relations]
An interleaved step relation interleaves thread-local reductions with
interference from concurrent threads captured as the rely relation
($R$).  It is defined thus:\vspace*{-10pt}

@@

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation thus:\vspace*{-10pt}

@@

\noindent The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to an RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
